Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging

### 2015-09-30 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:09:30	16:10:27	(Wed30.Sep)

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. We will express it in vector form as:
%
\begin{align*}
\f_\beta = \bmat{
f_0(\beta) \\
\vdots\\
f_{M-1}(\beta)
}
\end{align*}
%
where
%
\begin{align}
f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ is the extent of the window in time and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

@@

Throughout this article we have used LCA with variations of the Kaiser window function. However, we could instead have used a trigonometric window function:

\ref{trig_window_function} \ref{steering_bounds}

%
\begin{align}
f_m(\alpha) = e^{j\phi_\theta[m]}\Big[\alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\Big]\label{trig_window_function}
\end{align}
%
where $\phi_\theta[m] = \frac{2\pi}{\lambda}dm\sin\theta$ is the steering angle. For this window there is an analytic solution for the optimal value of $\alpha$ in the minimum variance sense. To derive it, we start by inserting the window function (\ref{trig_window_function}) into the beamformer equation (\ref{z}):
%
\begin{align}
\w\H\x &= \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
&= \alpha \a\T\x + (1-\alpha)\b\T\x
\end{align}
%
where $\a$ is the steering vector defined in (\ref{steering_vector}), and 
%
\begin{align}
\b = \text{diag}(\a)\cdot\begin{bmatrix}
     1 &
     \cos\left(\frac{2\pi}{M-1}\right) &
     \cos\left(\frac{2\pi 2}{M-1}\right) &
     \hdots &
     1
     \end{bmatrix}\T.
\end{align}
%
To attain unity gain in the look direction the weights must sum to one:
%
\begin{align}
\w\T\1 &= \big(\alpha\a - (1-\alpha)\b\big)\T\1 = 1
\end{align}
%
This is true for any value of $\alpha$ if $\a\T\1 = 1$ and $\b\T\1 = 1$. Hence, we can preserve the unity gain constraint as long as we normalize $\a$ and $\b$.

Now let $a=\a\T x$ and $b=\b\T x$. The beamformer output can then be written as:
%
\begin{align}
|\w\H\x|^2 &= \Big| \alpha a + (1-\alpha)b \Big|^2 \nn
&= \Big[\alpha a + (1-\alpha)b\Big] \Big[\alpha a^* + (1-\alpha)b^*\Big] \nn
&= \alpha^2 aa^* + \alpha a(1-\alpha)b^* + \alpha a^*(1-\alpha)b + (1-\alpha)^2bb^* \nn
&= \alpha^2 aa^* + \alpha(1-\alpha)(ab^* + a^*b) + (1-\alpha)^2bb^* \nn
&= \alpha^2 (aa^* - (ab^* + a^*b) - bb^*) + \alpha(ab^* + a^*b) + bb^* \nn
\end{align}
%
This is a convex function with a single minimum, which we obtain by differentiating with respect to $\alpha$ and setting equal to 0:
%
\begin{align}
\frac{\partial}{\partial\alpha} |\w\H\x|^2 
&= 2\alpha (aa^* - (ab^* + a^*b) - bb^*) + ab^* + a^*b = 0
\end{align}
%
which has the solution:
%
\begin{align}
\alpha &= \frac{ab^* + a^*b}{2(aa^* - (ab^* + a^*b) - bb^*)}
\end{align}
%
In this computation there are only 4 unique additions and multiplications. If we used this to analytically solve for $\alpha$, but perform the search for $\phi$, the computational complexity of LCA would be of O($8MN_\phi)$ instead of O($MN_\alpha N_\phi)$. The value for $\alpha$ would also be the optimal in the minimum variance sense.

>>>>

As will be demonstrated in upcoming sections, the LCA beamformer works very well with windows generated from the Kaiser-Bessel function. We will express it in vector form as:
%
\begin{align}
\f_\beta = \bmat{
f_0(\beta) \\
\vdots\\
f_{M-1}(\beta)
}\label{kaiser_window_vector}
\end{align}
%
where
%
\begin{align}
f_m(\beta) = \frac{I_0\left(\pi\beta\sqrt{1-\left(\frac{2m}{M-1}-1\right)^2}\right)}{I_0(\pi\beta)}\label{kaiser_window_element}
\end{align}
%
and $I_0$ is the zeroth order modified Bessel function of the first kind:
%
\begin{align}
I_0(x) = \sumb{a=0}{\infty} \left[ \frac{\left(\frac{x}{2}\right)^a}{a!} \right]^2.\label{modified_bessel_first_kind}
\end{align}
%
The Kaiser-Bessel window is near optimal in the sense of having its peak energy concentration around $\theta=0^\circ$, for a given time-bandwidth product specified with the parameter $\beta$:
%
\begin{align}
\beta = \frac{TB}{2},
\end{align}
%
where $T$ is the extent of the window in time and $B$ is its bandwidth. Adjusting $\beta$ changes the trade-off between mainlobe width and sidelobe level. When $\beta=0$ the window becomes rectangular, while at large values ($\beta>5$) the window converges to a Gaussian both in time and frequency. This class of windows is generally considered well suited for separating closely spaced sources with amplitudes of a high dynamic range~\cite{Harris1978}, they are easy to make, and they are optimal for any value of $\beta$.

@@

Throughout this article we have used LCA with variations of the Kaiser window function defined in (\ref{kaiser_window_element}), steered in . However, as shown in \Fig{image_quality} we can obtain similar performance using the trigonometric window function:

\ref{trig_window_function} \ref{steering_window}

%
\begin{align}
f_m(\alpha) = e^{j\phi_\theta[m]}\Big[\alpha - (1-\alpha)\cos\Big(\frac{2\pi m}{M-1}\Big)\Big]\label{trig_window_function}
\end{align}
%
(\ref{steering_vector})
where $\phi_\theta[m] = \frac{2\pi}{\lambda}dm\sin\theta$ is the steering angle. For this window there is an analytic solution for the optimal value of $\alpha$ in the minimum variance sense. To derive it, we start by inserting the window function (\ref{trig_window_function}) into the beamformer equation (\ref{z}):
%
\begin{align}
\w\H\x &= \sumb{m=0}{M-1} e^{j\phi_\theta[m]}\Big(\alpha - (1-\alpha)\cos\left(\frac{2\pi m}{M-1}\right)\Big) x_m^* \nn
&= \alpha \a\T\x + (1-\alpha)\b\T\x
\end{align}
%
where $\a$ is the steering vector defined in (\ref{steering_vector}), and 
%
\begin{align}
\b = \text{diag}(\a)\cdot\begin{bmatrix}
     1 &
     \cos\left(\frac{2\pi}{M-1}\right) &
     \cos\left(\frac{2\pi 2}{M-1}\right) &
     \hdots &
     1
     \end{bmatrix}\T.
\end{align}
%
To attain unity gain in the look direction the weights must sum to one:
%
\begin{align}
\w\T\1 &= \big(\alpha\a - (1-\alpha)\b\big)\T\1 = 1
\end{align}
%
This is true for any value of $\alpha$ if $\a\T\1 = 1$ and $\b\T\1 = 1$. Hence, we can preserve the unity gain constraint as long as we normalize $\a$ and $\b$.

Now let $a=\a\T x$ and $b=\b\T x$. The beamformer output can then be written as:
%
\begin{align}
|\w\H\x|^2 &= \Big| \alpha a + (1-\alpha)b \Big|^2 \nn
&= \Big[\alpha a + (1-\alpha)b\Big] \Big[\alpha a^* + (1-\alpha)b^*\Big] \nn
&= \alpha^2 aa^* + \alpha a(1-\alpha)b^* + \alpha a^*(1-\alpha)b + (1-\alpha)^2bb^* \nn
&= \alpha^2 aa^* + \alpha(1-\alpha)(ab^* + a^*b) + (1-\alpha)^2bb^* \nn
&= \alpha^2 (aa^* - (ab^* + a^*b) - bb^*) + \alpha(ab^* + a^*b) + bb^* \nn
\end{align}
%
This is a convex function with a single minimum, which we find by differentiating with respect to $\alpha$ and setting equal to 0:
%
\begin{align}
\frac{\partial}{\partial\alpha} |\w\H\x|^2 
&= 2\alpha (aa^* - (ab^* + a^*b) - bb^*) + ab^* + a^*b = 0,
\end{align}
%
which has the solution:
%
\begin{align}
\alpha &= \frac{ab^* + a^*b}{2(aa^* - (ab^* + a^*b) - bb^*)}.
\end{align}
%
In this computation there are only 4 unique additions and multiplications. If we used this to analytically solve for $\alpha$, but perform the search for $\phi$, the computational complexity of LCA would be of O($8MN_\phi)$ instead of O($MN_\alpha N_\phi)$. The value for $\alpha$ would also be the optimal in the minimum variance sense.
